---
title: "Intro to ggplot2 for R data visualizations"
format: gfm
editor: source
---

```{r settings, echo=FALSE, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, dpi=700)
```

## Motivation
- human brains are wired to process visual information quickly and recognize patterns
- well designed data visualizations can often bring a story to life and assist with further insights investigation
- ggplot2 is a popular R package for data visualization and a useful tool for analysts to have in their toolbox

## ggplot2 Overview

-   ggplot2 created by Hadley Wickham and part of the R tidyverse ecosystem
-   [package documentation and further learning resources](https://ggplot2.tidyverse.org/)
-   works best with tidy data input format (e.g. observations as rows and variables as columns)
-   pros
    -   open-source and free R package
    -   101 level knowledge can help analysts produce business value quickly
    -   used to produce publication worthy visualizations for tech and non-tech audiences
    -   highly flexible and customizable
    -   robust documentation, blog content, and community support (which means lots of training data for GenAI tools)
    -   high ROI on learning time invested; analysts can use the package across companies (vs paid tools which company A may use but not company B)
    -   [rich collection extension packages](https://exts.ggplot2.tidyverse.org/gallery/) (e.g. patchwork, gganimate, ggforce, ggtext, etc.)
    -   can be used to produce static and interactive visualizations for wide range of data types (e.g. time series, geospatial, etc.)
    -   proficient users can produce complex visualizations with a few lines of code (often faster than other data vis tools)
-   cons
    -   requires some basic R coding skills (R is known to have a slow / steep learning curve)
    -   advanced custom plots require deep knowledge of package nuances (which can be hard to remember); tools like Github Copilot reducing the cognitive load here
    -   data shaping/wrangling work needed at times
    -   plot outputs to Google Docs or Slides require partially manual workflow
    -   analyst-to-analyst shareability can be a challenge if the recipient is not familiar with R/ggplot2 (Google Sheets, GUI based tools, etc often more accessible across data teams)

## Key Components

-   **layers:** ggplot builds plots layer by layer in a logical order
-   **aesthetics mapping (aes):** instructions for how the data will map to a graphic (axis, colors, shapes, lines, etc)
-   **geometries (geoms)**: determines the graphic used to visually display the data
-   **facets**: used to split data into subplots based on 1 or more variables
-   **labs:** title, axis names, captions, legend titles, etc.
-   **themes**: used to reduce visual noise, add polish, and tweak formatting

## Packages

```{r packages}
# ggplot2 included in the tidyverse package
required_packages <- c('tidyverse', 'janitor', 'skimr')

### check for required packages and install them if they are not already installed
### not a best practice in general, but useful for folks getting started with R
for(p in required_packages) {
  if(!require(p,character.only = TRUE)) 
        install.packages(p, repos = "http://cran.us.r-project.org")
  library(p,character.only = TRUE)
}
```

## Practice Dataset

-   wine ratings and chemical properties of Vinho Verde red wine samples from the north of Portugal
-   UCI ML dataset (more details [here](https://archive.ics.uci.edu/ml/datasets/Wine+Quality))

```{r}
### read in the red wine dataset using csv location URL 
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-red.csv"
rw_df <- read_delim(url, delim=";")

### clean up column names automatically
rw_df <- rw_df %>% janitor::clean_names()

### Uncomment lines below output dataset info directly in the notebook
# url_2 <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality.names"
# read_lines(url_2)

### tails of the rating distribution are sparse
### re-categorize to make the data more balanced and easier to visualize
rw_df <- rw_df %>%
    ### SQL like case statement to create a new column
    mutate(quality_group = case_when(
            quality<=4 ~ "3 to 4",
            quality==5~ "5",
            quality==6 ~ "6",
            quality>6 ~ "7 to 8"),
           ### converts character column to a factor with level orderings
           ### level ordering used downstream for logical plot ordering
           quality_group = factor(quality_group, levels=c("3 to 4", "5", "6", "7 to 8")))

### view the dataset, observation count, column counts, data types, etc
glimpse(rw_df)
```

## Summary Stats

-   context building step to understand the dataset and its variables

```{r}
rw_df %>%
  # approach to quickly produce summary stats
  skimr::skim_without_charts() %>%
  # return the numeric subtable produced by skim above
  yank("numeric") %>%
  # no missing data and all variables have a complete rate of 1
  select(-n_missing, -complete_rate) %>%
  # rounds numeric columns to 2 decimal places
  mutate_if(is.numeric, ~round(., 2)) %>%
  # used to product output table
  # TODO: could change function reference here
  knitr::kable()
```

## Building the Layers Step by Step

1.  pipe (`%>%`) a dataset into ggplot

```{r}
### outputs blank plot object with no data mapping
rw_df %>%
  ggplot()
```

2.  set mapping for how the data will be visually represented

```{r}
# add x and y axis mapping to blank plot object (called aesthetics mapping)
rw_df %>%
  ggplot(aes(x = volatile_acidity, y = alcohol))
```

3.  specify a geom (e.g. graphic type) to visually represent the data

```{r}
# adds points to the plot object with x and y axis mapped
rw_df %>%
  ggplot(aes(x = volatile_acidity, y = alcohol)) +
  geom_point()
```

4.  adjust point size and alpha parameter to handle overplotting

```{r}
# darker region highlights the primary cluster of data
rw_df %>%
  ggplot(aes(x = volatile_acidity, y = alcohol)) +
  geom_point(size = 0.9, alpha = 0.4)
```

5.  use color to distinguish between different quality groups

```{r}
# set color to quality_group within the aesthetics mapping logic
rw_df %>%
  ggplot(aes(x = volatile_acidity, y = alcohol, color = quality_group)) +
  geom_point(size = 0.9, alpha = 0.4)
```

6.  use facets to split the data into subplots based on quality group

-   overplotting on above iteration making it hard to see differences between quality groups

```{r}
rw_df %>%
  ggplot(aes(x = volatile_acidity, y = alcohol, color = quality_group)) +
  # drop legend to reduce visual noise; facet headers provide the same info
  geom_point(size = 0.9, alpha = 0.4, show.legend = F) +
  theme(legend.position = "top") +
  # split the data into subplots based on quality group
  facet_wrap(. ~ quality_group, ncol = 2)
```

7.  add median vertical and horizontal lines to enhance group vs group comparison

```{r}
rw_df %>%
  ggplot(aes(x = volatile_acidity, y = alcohol, color = quality_group)) +
  geom_point(size = 0.9, alpha = 0.4, show.legend = F) +
  ### dynamically add median reference lines
  geom_vline(aes(xintercept = median(volatile_acidity)), 
             linetype = "dashed", color = "grey40") +
  geom_hline(aes(yintercept = median(alcohol)), 
                 linetype = "dashed", color = "grey40") +
  facet_wrap(. ~ quality_group, ncol = 2)
```

8.  add titling and finalize styling

```{r}
### assign plot object to a variable (used downstream)
p1 <- rw_df %>%
  ggplot(aes(x = volatile_acidity, y = alcohol, color = quality_group)) +
  ### note linetype aes mapping "hack" to surface the median legend title
  geom_vline(aes(xintercept = median(volatile_acidity), 
                 linetype = "Median"), color = "grey40",
                 show.legend = F) +
  geom_hline(aes(yintercept = median(alcohol), 
                 linetype = "Median"), color = "grey40") +
  geom_point(size = 0.9, alpha = 0.4, show.legend = F) +
  facet_wrap(. ~ quality_group, ncol = 2) +
  labs(title = "Red Wine Alcohol and Volatile Acidity by Quality Group",
       subtitle = "top rated wines tend to have lower volatile acidity and higher alcohol %",
       linetype = "",
       x = "Volatile Acidity",
       y = "Alcohol %")
p1
```

9.  save the plot to an image file

```{r}
### saved to working directory
ggsave("red_wine_quality_group_alcohol_and_va.png", p1, width = 10, height = 6)
```


Next step bookmark
- create 3 additional data vis examples with simple code and strong visuals
- TBD on programming and scaleable visualization sections


## Outline notes

-   ggplot pros and cons
-   ggplot overview
-   getting started examples
-   faceting
-   styling and customization
-   programmatic visualization
-   scaleable visualization
-   additional resources
