---
title: "Intro to ggplot2 for R data visualizations"
format: gfm
editor: source
---

```{r settings echo=FALSE, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

```{r packages}
required_packages <- c('tidyverse', 'rattle')

### checks for required packages and install them if they are not already installed
### not a best practice in general, but useful for folks getting started with R
for(p in required_packages) {
  if(!require(p,character.only = TRUE)) 
        install.packages(p, repos = "http://cran.us.r-project.org")
  library(p,character.only = TRUE)
}
```

## Outline

-   ggplot pros and cons
-   ggplot overview
-   getting started examples
-   faceting
-   styling and customization
-   programmatic visualization
-   scaleable visualization
-   additional resources

## Pros and Cons to ggplot2

+--------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------+
| Pros                                                                                                                                       | Cons                                                                   |
+--------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------+
| -   open-source and free R package                                                                                                         | -   requires some basic R coding skills                                |
| -   101 level knowledge can help analysts produce business value                                                                           | -   highly customize plots require deep knowledge of package nuances   |
| -   can produce publication worthy visualizations for tech and non-tech audiences                                                          | -   data shaping/wrangling work needed at times                        |
| -   highly flexible and customizable                                                                                                       | -   plot outputs to Google Docs or Slides often require manual process |
| -   robust documentation, blog content, and community support (which means lots of training data for GenAI tools)                          |                                                                        |
| -   high ROI on learning time given analysts can use the package across companies (vs tool paid which company A may use but not company B) |                                                                        |
+--------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------+

## Ggplot2 Overview

-   created by Hadley Wickham and part of the broader tidyverse R ecosystem
-   works best with tidy data format (observations as rows and variables as columns)
-   key concepts:
    -   **layers:** ggplot builds plots layer by layer in a logical order
    -   **aesthetics mapping**: instructions for how the data will map to a graphic (axis, colors, shapes, lines, etc)
    -   **geometries (geoms)**: determines the graphic used to visually display the data
    -   **facets**:
    -   **themes**: used to reduce visual noise, add polish, and tweak formatting.

## Getting Started Examples

-   UCI ML dataset: `mlbench::WineQualityRed`
-   wine ratings and chemical properties of Vinho Verde rednwine samples from the north of Portugal
-   more details [here](https://archive.ics.uci.edu/ml/datasets/Wine+Quality)

```{r}
# where the ratings? what is type? Mighty need another dataset
wine
```


