---
title: "Intro to ggplot2 for R data visualizations"
format: gfm
editor: source
---

### Pros and Cons to ggplot2

+--------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------+
| Pros                                                                                                                                       | Cons                               |
+--------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------+
| -   open-source and free R package                                                                                                         | -   \<\<\<Next step bookmark\>\>\> |
| -   101 level knowledge can help analysts produce business value                                                                           |                                    |
| -   can produce publication worthy visualizations for tech and non-tech audiences                                                          |                                    |
| -   highly flexible and customizable                                                                                                       |                                    |
| -   robust documentation, blog content, and community support (which means lots of training data for GenAI tools)                          |                                    |
| -   high ROI on learning time given analysts can use the package across companies (vs tool paid which company A may use but not company B) |                                    |
+--------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------+

```{r settings echo=FALSE, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

```{r packages}
required_packages <- c('tidyverse')

### check for required packages and install them if they are not already installed
### not a best practice in general, but useful for folks getting started with R
for(p in required_packages) {
  if(!require(p,character.only = TRUE)) 
        install.packages(p, repos = "http://cran.us.r-project.org")
  library(p,character.only = TRUE)
}
```

## Outline

-   ggplot pros and cons
-   ggplot overview
-   getting started examples
-   faceting
-   styling and customization
-   programmatic visualization
-   scaleable visualization
